An Analysis of Performance Bottlenecks in MRI Pre-Processing

### 2024-03-25 paper/paper.tex
Finished first draft.

\begin{abstract}
	% TODO																
\end{abstract}

@@

%  GPUs
\MD{TODO GPUs / Accelerators}

@@

\subsubsection{ANTS registrationSyN}
% Spatial Normalization
% SyN Registration: https://doi.org/10.1016/j.media.2007.06.004
% TODO Mention the option for single and double precision.

\subsubsection{ANTS brainExtraction}
% ANTS brainExtraction
% 1. Truncate image intensity
% 2. N4 bias correction: https://doi.org/10.1109/TMI.2010.2046908
% 3. SyN Registration: : https://doi.org/10.1016/j.media.2007.06.004
% 4. Warp prior to image using affine
% 5. Atropos: https://pubmed.ncbi.nlm.nih.gov/21373993/
% TODO Mention the option for single and double precision.
% TODO OASIS template citation

\subsubsection{FSL FAST}
% DOI: https://doi.org/10.1109/42.906424
% Segmentation

\subsubsection{FreeSurfer recon-all}
% Surface Reconstruction: https://doi.org/10.1006/nimg.1998.0395
% Refer to the fMRIPrep specific 3 stages workflow.

\subsubsection{FSL FLIRT}
% Registration
% FSL FLIRT: https://doi.org/10.1006/nimg.2002.1132, https://doi.org/10.1016/S1361-8415(01)00036-6, https://doi.org/10.1016/j.neuroimage.2009.06.060

\subsubsection{FSL MCFLIRT}
% Head-Motion Correction
% FSL MCFLIRT: https://doi.org/10.1006/nimg.2002.1132

\subsubsection{fMRIPrep}
\MD{TODO import fMRIPrep citation boilerplate from "./fmriprep-citations"}

@@

The Intel VTune profiler solves these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level. However, common profiling challenges remain. Profilers require source code to be compiled with Debug Symbols to report useful information about the different functions and modules. Moreover, different input data can substantially vary the analysis due to conditional branching in the application and convergence thresholds. We discuss our approach to these challenges in the next section.
\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for refereces}

@@

The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load. L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache. Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively. The store bound occurs when a store operation stall the CPU. Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory boudnd characterization~\cite{Intel2006-lc}. We use the convention from~\cite{Kukunas2015-jd} for the equations.

@@

\subsection{Makespan: A performance overview}
Figure~\ref{fig:makespan-1thread} depicts the total time elapsed from the start to the end of each pipeline; i.e. makspan. There is a significant difference between the makespan of different applications. This is expected as the pipelines perform different tasks. The standard deviation is nearly null for all pipelines, except for FreeSurfer recon-all which as a small variation. Overall, this shows that the input data does not substantially affect the makespan of those pipelines.

@@

%%%%%%%%%% # Linear Interpolation
% Theoretical improvements using reduced precision. If time permits, toy experiment with ITK single precision data type at different locations.
% 1. Winograd algorithm for fast convolution.
% - It could be used to speed up the current pipeline. Tradeoff with precision? How does it compare to reduced precision?
% 2. Improvement to interpolation would help the current algorithms, as well as the DL algorithm using CNN.
\subsection{Importance of Linear Interpolation}
\MD{TODO}

%%%%%%%%%% # Single vs. Double Precision
% 1. Not as simple as using different data type. It could affect
% - Convergence
% - Accuracy
% - Runtime (maybe even slower)
% 1.5 Convergence is the sane in our case (same number of iterations).
% 2. Why is it slower in our case?
% -? Possibly less optimized code for single precision.
% -? Possibly memory placement
% -? ...
% 3. 
\subsection{ANTS: A tale of precision}
\MD{TODO + change title}

@@

@@

\section{Conclusion}
\MD{TODO}

@@

Our profiling results from VTune are publicly available at:
\\\href{URL}{\MD{TODO add Zenodo link}}

@@

\section*{Acknowledgement}
\MD{TODO}
% NSERC PGS-D
% Concordia Bursary (find name)
% Any from others?
% ...

@@

\newpage
\onecolumn
\section*{Supplementary Materials}
\begin{figure*}
	\centering
	\includegraphics{figures/global-zipf_law.png}
	\caption{Distribution of the functions' CPU time compare to a Zipf's law distribution with $\alpha=1.8$. The y-axis shows the average CPU time for a function. The x-axis shows the percentage of functions ordered by decreasing CPU time. The data includes all functions from all pipelines.}
	\label{sup-fig:zips-law}
\end{figure*}

\label{sec:supplementary}

>>>>

\begin{abstract}
	MRI pre-processing is a critical step for neuroimaging analysis. However, the computational cost of MRI pre-processing pipelines is a major bottleneck for large cohort studies and some clinical applications. While HPCs and, more recently, Deep Learning have been adopted to accelerate the computations, these techniques require costly hardware and are not accessible to all researchers. Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance. Using Intel VTune profiler, we characterize the bottlenecks of several commonly used MRI-preprocessing pipelines from the ANTs, FSL, and FreeSurfer toolboxes. First, we confirmed that few functions contribute to most of the CPU time. Second, we found that linear interpolation is the main bottleneck for the applications. Data access is also  a substantial bottleneck for the pipelines. Last, we found a bug in the ITK library that impact the performance of ANTs pipeline in single-precision and a potential issue with the OpenMP scaling in FreeSurfer recon-all. Our results provide a reference for future efforts to optimize MRI pre-processing workflows.
\end{abstract}

@@

@@

We use the default parameters for fMRIPrep and most of the sub-workflows. However, for FreeSurfer recon-all, we mimic the fMRIPrep specific 3 stages workflow.

We profile both the single and double precision version for ANTs registrationSyN and ANTs brainExtraction. ANTs uses ITK templates to implement the single precision alternative.

@@

The Intel VTune profiler solves these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level. However, common profiling challenges remain. Profilers require source code to be compiled with Debug Symbols to report useful information about the different functions and modules. Moreover, different input data can substantially vary the analysis due to conditional branching in the application and convergence thresholds. We discuss our approach to these challenges in the next section.
\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for references}

@@

The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load. L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache. Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively. The store bound occurs when a store operation stall the CPU. Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc}. We use the convention from~\cite{Kukunas2015-jd} for the equations.

@@

\subsection{Makespan: A performance overview}
Figure~\ref{fig:makespan-1thread} depicts the total time elapsed from the start to the end of each pipeline; i.e. makespan. There is a significant difference between the makespan of different applications. This is expected as the pipelines perform different tasks. The standard deviation is nearly null for all pipelines, except for FreeSurfer recon-all which as a small variation. Overall, this shows that the input data does not substantially affect the makespan of those pipelines.

@@

\subsection{Importance of Linear Interpolation}
Our results show that linear interpolation the primary bottleneck for all pipelines, except FreeSurfer recon-all. While each interpolation have low computational cost, the large amount of operations make it a significant bottleneck.

There is an extensive literature on interpolation. Unfortunately, large amount of optimization techniques are not applicable to MRI pre-processing pipelines, because of the rotation during registration which requires dynamic kernels for interpolation. Moreover, the high-resolution of the data and iterative refinement of the registration require a large number of interpolations to be performed.

The authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation. The review algorithm trade-off accuracy for a lower computational cost. We think it would be interesting for future work to explore these different alternatives to accelerate interpolation within the pipelines. However, special attention will be required to balance the accuracy and the performance of the algorithm.

Similarly, we propose future efforts to explore reduced- or mixed-precision techniques to accelerate interpolation. Optimizing the data format used for interpolation would reduce the computational cost, memory footprint, and storage requirements.

Improvements to interpolation in 3D MRI imaging would have a large impact. It would also benefit other fields of studies using interpolation for high-resolution imaging with iterative refinement, such as climate simulation, astronomical imaging, or 3D image reconstruction.

\subsection{ANTS: A tale of precision}
From our results, we observe a slow-down for the single precision version of ANTs pipelines. A single iteration takes longer to execute when using single-precision compared to double precision. After a further analysis of the ANTs registrationSyN single-precision pipeline, we found that the function\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} contributing most of the CPU time does computations in double precision. Therefore, the pipeline wastes CPU cycles to convert the input data to double in ITK, and potentially to cast to back to float in ANTS. We think this bug is the cause of the slow-down. We tried recompiling a version of ITK to fix this issue, but we were unsuccessful due to the size and complexity of the code base. This example shows that the use of reduced precision is not trivial and require a deep understanding of the code base. Moreover, benchmark should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.

@@

\subsection{VTune results finalization}
Using VTune to collect profiling results does not add a large overhead in term of time and resources utilization. However, a finalization step is required to query the data and generate a VTune report. We found this step to be both time consuming and resource intensive. For example, finalizing the results from a single FreeSurfer recon-all execution requires up to \SI{2}{\tera\byte} of RAM, while the developers recommend to use between \SI{8}{\giga\byte} to \SI{16}{\giga\byte} of RAM to run the pipeline. This significant difference in resource requirement creates a challenge to profile pipeline with long runtime, due to the potential difficulty to obtain access to compute nodes with enough RAM. This can be further challenging for researchers in early career or in developing countries, which might not have access to large HPC infrastructure.
One way to mitigate this challenge is to lower the sampling rate during the profiling. However, to reach a similar level of memory consumption, the sampling rate must be significantly reduced. This would lead to a potential loss of information in the profiling results.

@@

\section{Conclusion}
In this paper, we profiled several MRI pre-processing applications: ANTS brainExtraction, ANTS registration, FSL FAST, FSL FLIRT, FSL MCFLIRT, FreeSurfer recon-all, and fMRIPrep. We confirmed our hypothesis that few functions contribute to a large amount of CPU time. We found that most applications suffered from memory bound. Linear interpolation was the primary bottleneck for these applications. We discovered a bug in ITK which lead the implementation of ANTS registration in single-precision to use float for its computation. We also discovered a potential bug in FreeSurfer recon-all which limits the benefit from multi-threading with OpenMP. Last, we discuss the challenge of profiling long running applications with VTune, due to computational resource requirements to obtain results.

We suggest that future effort focus on reduced precision techniques, which could reduce memory bound and accelerate interpolation computation. We note that the use of reduced precision is not trivial, as seen with the ANTS bug. Therefore, careful attention would be required to find a balance between performance and accuracy, while ensuring bugs are not introduced. We also suggest that future work should study the different configuration of OpenMP scheduling in FreeSurfer recon-all to improve the parallel efficiency. This would lead to faster runtime time and higher CPU usage ration in HPC allocations.

We hope that this work serves as a reference for future work to optimize MRI pre-processing pipelines.

@@

Our profiling results from VTune are publicly available at:
\\\href{URL}{\MD{TODO add Zenodo link}}
% Waiting on data transfer from Compute Canada

@@

\section*{Acknowledgement}
Mathieu Dugr'e was jointly funded by the Natural Sciences and Engineering Research Council of Canada (NSERC) through the Postgraduate Scholarship-Doctoral (PGS-D) program and the Graduate Doctoral Fellowship Award from Concordia University.
\MD{Q: Any other?}

@@

\newpage
\onecolumn
\section*{Supplementary Materials}
\MD{Not sure if these add value?}
\label{sec:supplementary}

\begin{figure*}[h]
	\centering
	\includegraphics{figures/global-zipf_law.png}
	\caption{Distribution of the functions' CPU time compare to a Zipf's law distribution with $\alpha=1.8$. The y-axis shows the average CPU time for a function. The x-axis shows the percentage of functions ordered by decreasing CPU time. The data includes all functions from all pipelines.}
	\label{sup-fig:zips-law}
\end{figure*}
